Attribution of undesirable character traits, rather than trait-based dehumanization, predicts punishment decisions

### 2023-06-26 manuscript.Rmd
Update manuscript.Rmd

The construct of dehumanization is often invoked to explain intergroup harm. In key evidence for this claim, @Bastian2013 reported that the extent to which participants dehumanized criminals by denying them uniquely human character traits such as refinement, rationality and morality predicted the severity of the punishment participants endorsed for them. We revisit this influential finding across six highly powered and preregistered studies. In Studies 1A and 1B, we conceptually replicate Bastian and colleagues' effects, demonstrating that our method is sensitive to detecting relationships between trait-based dehumanization and punishment should they occur. In subsequent studies, we investigate whether the apparent relationship between trait-based dehumanization and punishment is driven by the desirability of the traits incorporated into the stimulus set, their perceived humanness, or both. Whereas Bastian et al. asked participants to rate the extent to which criminals possessed uniquely human traits that were socially desirable, we asked participants to rate the extent to which criminals possessed uniquely human traits that were both socially desirable (e.g., cultured, civilised) and socially undesirable (e.g., arrogant, bitter). Correlational (studies 2A and 2B) and experimental (studies 3A and 3B) evidence converge on the conclusion that apparent evidence for the relationship between trait-based dehumanization and punishment is better explained by the extent to which participants attribute socially desirable attributes to criminals rather than the extent to which they attribute uniquely human attributes to criminals. Taken together, these studies cast doubt on the hypothesised causal relationship between trait-based dehumanization and harm, at least in this context.

*Keywords*: dehumanization, harm, intergroup bias, punishment, social cognition, trait attribution

@@

One way to de-confound evidence for trait-based dehumanization from evidence of negative evaluation is to ask participants to rate the target group on traits that are uniquely human but socially undesirable [@Over2020a; -@Over2020b]. Previous research has established that undesirable character traits such as jealousy, spite and bitterness are considered unique to humans and socially undesirable [@Enock2021a]. Enock and colleagues [-@Enock2021a] found that whereas participants tend to attribute socially desirable human traits more strongly to the ingroup, they tend to attribute socially undesirable traits more strongly to the outgroup. This result held across three intergroup contexts [see also @Enock2021b; @Enock2022]. Enock and colleagues conclude that apparent evidence for trait-based dehumanization may be better explained by intergroup preference. However, it is not yet clear how the attribution of uniquely human character traits relates to harm.

We revisit the hypothesised causal relationship between trait-based dehumanization and harm in the context of endorsing harsh punishment for criminals. In Studies 1A and 1B, we seek to conceptually replicate the key findings of Bastian and colleagues [-@Bastian2013] suggesting that the extent to which participants animalistically (Study 1A) and mechanistically (Study 1B) dehumanize criminals predicts the severity of the punishment participants endorse for them. In Studies 2A and 2B, we adopt a similar design but incorporate socially undesirable traits into our stimulus set. This addition to the design allows us to investigate whether trait-based dehumanization predicts the severity of punishment, undesirable trait attribution predicts the severity of punishment, or both predict the severity of punishment. In Studies 3A and 3B, we seek to investigate a similar question using an experimental design. We present participants with vignettes in which criminals are described using character traits that differ in how socially desirable they are and in whether or not they are unique to humans and measure how these varying descriptions influence participants' parole decisions. This allows us to directly measure whether there is a causal relationship between trait-based dehumanization and punishment.

@@

All studies received ethical approval from the Psychology Departmental Ethics Committee University of York. All data collection occurred online, and the studies were created and administered using Qualtrics (https://www.qualtrics.com). Participants were recruited through the online platform Prolific (https://www.prolific.co), with an independent sample recruited for each study. Informed consent was obtained at the start of each session according to approved ethical guidelines. Inclusion criteria for each study included adult participants fluent in English who had never been to prison for committing a crime and had a Prolific approval rating of at least 90% (95% for Studies 3A and 3B). Increases in Prolific's recommended rate of compensation for participation during data collection meant the reward ranged from approximately £7 per hour in Studies 1A and 1B to approximately £8 in the other four studies. Assumption testing and analyses were conducted using *SPSS* and *RStudio*. All studies were pre-registered on AsPredicted.com before commencing data collection. Links to pre-registration documents, data files, files for computational replication on GitHub, and supplementary materials including the stimuli used for each study can be found at: https://osf.io/d4cvp/?view_only=ccf139206e59471b979f53708595ef05.

@@

A power analysis using G\*Power indicated that a sample size of 135 would allow us to detect a medium effect size ($\eta_p^2$ = .09) with a $2 \times 2$ factorial design, an alpha of .05 and a power of .95. To counterbalance the sample equally, a sample of 136 participants was collected, of whom `r gen_data3A[1]` identified as female, `r gen_data3A[2]` as male, `r gen_data3A[3]` as non-binary, and `r gen_data3A[4]` who preferred not to indicate their gender identity.  Ages ranged from `r min_age3A` to `r max_age3A` (*M* = `r round(age_mean3A, digits = 1)`, *SD* = `r round(age_sd3A, digits = 2)`).  All participants were adults fluent in English who had never been to prison for committing a crime.  Due to a noticeable increase in failed attention checks during pilot data collection, the minimal approval rating on Prolific was raised from 90% to 95%.  Despite this, data from 46 participants were omitted and replaced due to failed attention checks. Three participants were mistakenly recruited after the intended sample size had been met, and thus, their data were excluded from analyses^[Including data submitted by excess participants in analyses yielded the same results as those reported.]. Participation took an average of seven minutes.

@@

**Attention check**. An additional paragraph was included, largely similar to the other four paragraphs, describing a criminal named Charlie. However, in the middle of the paragraph, the following sentence was included: '*This paragraph is an attention check: please move the slider all the way to Strongly Disagree on the left-hand side*'. Data submitted by any participants who did not respond within 20 points of the instructed end of the 100-point scale were omitted and replaced.

>>>>

The construct of dehumanization is often invoked to explain intergroup harm. In key evidence for this claim, @Bastian2013 reported that the extent to which participants dehumanized criminals by denying them uniquely human character traits such as refinement, rationality and morality predicted the severity of the punishment endorsed for them. We revisit this influential finding across six highly powered and preregistered studies. In Studies 1A and 1B, we conceptually replicate Bastian and colleagues' effects, demonstrating that our method is sensitive to detecting relationships between trait-based dehumanization and punishment should they occur. In subsequent studies, we investigate whether the apparent relationship between trait-based dehumanization and punishment is driven by the desirability of the traits incorporated into the stimulus set, their perceived humanness, or both. Whereas Bastian et al. asked participants to rate the extent to which criminals possessed uniquely human traits that were socially desirable, we asked participants to rate the extent to which criminals possessed uniquely human traits that were both socially desirable (e.g., cultured, civilised) and socially undesirable (e.g., arrogant, bitter). Correlational (studies 2A and 2B) and experimental (studies 3A and 3B) evidence converge on the conclusion that apparent evidence for the relationship between trait-based dehumanization and punishment is better explained by the extent to which participants attribute socially desirable attributes to criminals rather than the extent to which they attribute uniquely human attributes to criminals. Taken together, these studies cast doubt on the hypothesised causal relationship between trait-based dehumanization and harm, at least in this context.

*Keywords*: dehumanization, harm, intergroup bias, social cognition, trait attribution

@@

One way to de-confound evidence for trait-based dehumanization from evidence of negative evaluation is to ask participants to rate the target group on traits that are uniquely human but socially undesirable [@Over2020a; -@Over2020b]. Previous research has established that undesirable character traits such as jealousy, spite and bitterness are considered unique to humans and socially undesirable [@Enock2021a]. Enock and colleagues [-@Enock2021a] found that participants tend to attribute socially desirable human traits more strongly to the ingroup, and socially undesirable traits more strongly to the outgroup. This result held across three intergroup contexts [see also @Enock2021b; @Enock2022]. Enock and colleagues conclude that apparent evidence for trait-based dehumanization may be better explained by intergroup preference. However, it is not yet clear how the attribution of uniquely human character traits relates to harm.

We revisit the hypothesised causal relationship between trait-based dehumanization and harm in the context of endorsing harsh punishment for criminals. In Studies 1A and 1B, we seek to conceptually replicate the key findings of Bastian and colleagues [-@Bastian2013] suggesting that the extent to which participants animalistically (Study 1A) and mechanistically (Study 1B) dehumanize criminals predicts the severity of the punishment participants endorse for them. In Studies 2A and 2B, we adopt a similar design but incorporate socially undesirable traits into our stimulus set. This addition to the design allows us to investigate whether trait-based dehumanization predicts the severity of punishment, undesirable trait attribution predicts the severity of punishment, or both predict the severity of punishment. In Studies 3A and 3B, we seek to investigate a similar question using an experimental design. We present participants with vignettes in which criminals are described using character traits that differ in how socially desirable they are and whether or not they are unique to humans. We also measure how these varying descriptions influence participants' parole decisions. This allows us to directly measure whether there is a causal relationship between trait-based dehumanization and punishment.

@@

All studies received ethical approval from the Psychology Departmental Ethics Committee University of York. All data collection occurred online, and the studies were created and administered using Qualtrics (https://www.qualtrics.com). Participants were recruited through the online platform Prolific (https://www.prolific.co), with an independent sample recruited for each study. Informed consent was obtained at the start of each session according to approved ethical guidelines. Inclusion criteria for each study included adult participants fluent in English who had never been to prison for committing a crime and had a Prolific approval rating of at least 90% (95% for Studies 3A and 3B). Increases in Prolific's recommended rate of compensation for participation during data collection meant the reward ranged from approximately £7 per hour in Studies 1A and 1B to approximately £8 in the other four studies. Assumption testing and analyses were conducted using *SPSS* and *RStudio*. All studies were pre-registered on AsPredicted.com before commencing data collection. Links to pre-registration documents, data files [@Brennan2023], a fully computationally reproducible version of the manuscript, and supplementary materials including the stimuli used for each study can be found at: https://osf.io/d4cvp/?view_only=ccf139206e59471b979f53708595ef05.

@@

A power analysis using G\*Power, with effect size specification as in SPSS, indicated that a sample size of 135 would allow us to detect a medium effect size ($\eta_p^2$ = .09) with a $2 \times 2$ factorial, repeated measures design, an alpha of .05, and a power of .95. To counterbalance the sample equally, a sample of 136 participants was collected, of whom `r gen_data3A[1]` identified as female, `r gen_data3A[2]` as male, `r gen_data3A[3]` as non-binary, and `r gen_data3A[4]` who preferred not to indicate their gender identity.  Ages ranged from `r min_age3A` to `r max_age3A` (*M* = `r round(age_mean3A, digits = 1)`, *SD* = `r round(age_sd3A, digits = 2)`).  All participants were adults fluent in English who had never been to prison for committing a crime.  Due to a noticeable increase in failed attention checks during pilot data collection, the minimal approval rating on Prolific was raised from 90% to 95%.  Despite this, data from 46 participants were omitted and replaced due to failed attention checks. Three participants were mistakenly recruited after the intended sample size had been met, and thus, their data were excluded from analyses^[Including data submitted by excess participants in analyses yielded the same results as those reported.]. Participation took an average of seven minutes.

@@

**Attention check**. An additional paragraph describing a criminal named Charlie was included, largely similar to the other four paragraphs. However, in the middle of the paragraph, the following sentence was included: '*This paragraph is an attention check: please move the slider all the way to Strongly Disagree on the left-hand side*'. Data submitted by any participants who did not respond within 20 points of the instructed end of the 100-point scale were omitted and replaced.
